Relationships in the extremes and their influence on competition and coexistence

### 2022-03-22 paper/Paper.Rnw
continued wworking on Discussion

%***DAN: Go through and make sure all citation callouts list papers in date order, oldest to newest,
%except when multiple pubs by the same first author are listed they are all together.
%***JASMIN: Done

@@

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is the invader.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
Though some readers may find it slightly confusing to simultaneously use the variable 
$E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would 
be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like
all operators, $\E(\cdot)$ is rendered upright.

@@

We produced and applied a new extension to MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables, and our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association
such as ATAs.

%***DAN: Dan made it to here in editing and has to stop until after spring break

%The example case exapanding upon coexistence decompositions of the \cite{Descamps_2005} chemostat experiment support our earlier findings that ATA contributions are not only non-zero, but are non-neglible and contribute meaningfully to coexistence.

%Our expansion allows for the parsing of storage effects into contribution from a pure symmetrical correlation and structures beyond symmetric correlation. Previous studies report storage effect as a contribution to coexistence of EC covariance; yet compute it as the contribution of the EC relationship (covariance and beyond; symmetric and asymmetric).

%\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence or absence of ATAs 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – GWR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values, ATA exclusion occurs where ATAs are the 
driving mechanism of the failure to coexist (exclusion) and the lack of ATAs would result in 
a positive GWR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
One can imagine a species whose coexistence with others is conditional on the asymmetric relationship between drivers of coexistence and whose populations may become excluded when ATAs erode. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

%\underline{future work: study ATAs in environmental variables and how theyre changing - tie in with specialist}

Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of the magnitude of ATA contribution and environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in [the marginal/ univariate distributions of] extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

\underline{multispecies models}
%Need a para about multi-species models - how would things change? One difference is there are more different
%things that might have asymmetric tail association patterns. See the section "General Theory" of Ellner et al 2019 - might
%just want write a short para saying it seems like it should be possible to generalize what we've done along those lines.

@@

The chances of ATAs being important for the coexistence of species increases as more variables are included in the system, 
either through the addition of more species or more drivers (beyond just E and C), or both. 
When multiple species are included in the system, the interaction between different species' E's and C's introduces more relationships 
to consider that may or may not be asymmetrically related and could increase the relative importance of ATA contribution to the species' coexistence. 
Likewise, coexistence between species can also depend on more than two drivers \citep{Ellner_2019}. 
The storage effect decomposition presented here is generalizable to coexistence decompositions of multiple variables.
Following the set up and notation of \cite{Ellner_2019}'s \emph{covariance decomposition}, 
ATA contributions and correlation \emph{per se} can be decomposed from the \emph{associations} between two variables
among many from the appropriate term with parenthesized superscripts with the # symbol denoting and separating 
variables that are varying independently from the variables that are covarying whose relationship we are interested in decomposing. 
For example, the term $\Delta^{(XY#Z)}$ could be decomposed into $\Delta^{[XY#Z]} + \Delta^{[X \nat Y#Z]}$. 
As more terms and species are introduced into a system, more relationships between variables must be evaluated,
more of those relationships may be asymmetrically tail associated, and ATAs may contribute more to the maintenance of 
biodiversity in that system. 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,

\underline{caveats and future directions}
%Realism - we usede simple/extreme ATA, simple model and simple lab system, futurw work needed to assess 
%how mcuh ATA are important in complex systems, field systems
%what systems? what env vars?
%Ellner et al (2016) points out that coexistence is also mediated by other factors that competition: shared enemies, 
%mutualist, facilitation etc. (these are the exact factors Ellner lists), so future work could explore 
%the importance of ATA for coexistence mediated by these factors.
%Something about spatial mechanisms of coexistence? We considered only mechanisms that applied in a model with
%solely temporal variation, but there are also spatial mechanisms. Need to learn about those and think about
%how to say something about how future work may be able to explore the importance of ATA in that context?

@@

\begin{figure}
\includegraphics[width=.75\textwidth]{../results_figs/fig2.pdf}
\caption{Our simulated lottery-model populations influenced by asymmetric tail associations (ATAs) recovered 
from low densities less often and 
spent more time in a state of one-species dominance, demonstrating that ATAs can influence aspects of 
species coexistence. Panel (a) shows species 1 from the lottery 
model subject to environmental noise with symmetric tail associations (Fig. \ref{fig:noise}b). Panel (b) shows species 1 from the 
same model with the same parameters, but subject to noise with left-tail association (Fig. \ref{fig:noise}a).
Both simulations were run for \Sexpr{M} time steps, though only the first $1000$ are shown. Periods of 
``noticeable coexistence'' of the species, defined to be periods when each species comprises at least $5\%$ of 
total abundance, are marked in grey. The average duration of periods of noticeable coexistence 
was \Sexpr{fig2$coMets$Sym$co} time steps for panel (a) and \Sexpr{fig2$coMets$Asy$co} for panel (b), over the full length of the simulations; 
whereas the average duration of periods of
one-species dominance was \Sexpr{fig2$coMets$Sym$dom} time steps for panel (a) and \Sexpr{fig2$coMets$Asy$dom} for (b). Thus, while periods of 
noticeable coexistence were of similar duration for the two scenarios, periods of one-species dominance were much longer,
on average, for the ATA case. Parameters were $\mu_1 = \Sexpr{fig2$params[1]$,
$\mu_2 = \Sexpr{fig2$params[2]}$, $\sigma = \Sexpr{fig2$params[3]}$, and $\delta = \Sexpr{fig2$params[4]}$. 
See Introduction and Fig. \ref{fig:noise} caption for model details. 
\label{fig:pedagog}}
\end{figure}

>>>>

@@

The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``growth when rare'' [GWR] of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition on either species have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.
Though some readers may find it slightly confusing to simultaneously use the variable 
$E_i$ and the operator $\E(\cdot)$, both of these notational choices are standard, so it would 
be more confusing to use alternative notation. Like all variables, $E_i$ is rendered in italic; like
all operators, $\E(\cdot)$ is rendered upright.

@@

We produced and applied a new extension of MCT that 
quantifies the impact of ATAs on species coexistence and the maintenance of biodiversity. 
Building upon a recent, simulation-based approach to MCT \citep{Ellner_2016, Ellner_2019},
we decomposed storage effects into two new mechanisms, the contribution of ATAs to coexistence,
and the contribution of $EC$ correlation \emph{per se}.
Applications of our extended theory to the lottery model and to models based on a laboratory 
system of competing diatoms demonstrated that  
ATA contributions to storage effects and to species coexistence and the maintenance of biodiversity
can often be comparable in magnitude to the contributions of previously recognized mechanisms.
Our results highlight that common methods of correlation and covariance fail to include 
important information on the structure of relationships between variables. Our results add
coexistence and biodiversity maintenance to the list of ecological theories and phenomena on which 
ATAs have been shown to have an important impact \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of $EC$ \emph{covariance} to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019},
our results suggest that storage effects should be redefined to constitute the 
contribution of \emph{associations} between $E$ and $C$, including both ATA effects and
correlation \emph{per se}; the prior terminology, \emph{covariance}, is typically construed 
as reflecting the linear relationship between variables, ignoring complexities of the association
such as ATAs.

%ATA specialists
The contributions we have documented of ATAs to species coexistence open the possibility that
``ATA specialists'' may exist in some contexts. If a stronger competitor is present across a
landscape and the GWR of a weaker competitor %***DAN: Dan is working right here.

%Not only is the contribution of ATA non-neglible but in some cases the presence or absence of ATAs 
%can mediate the result of competition between species, i.e., coexistence or extirpation. 
%For some parameters, the contribution of ATAs was the driving mechanism of 
%coexistence -- GWR would be negative without it -- deeming the scenario a case of 
%ATA rescue. For other parameter values, ATA exclusion occurs where ATAs are the 
%driving mechanism of the failure to coexist (exclusion) and the lack of ATAs would result in 
%a positive GWR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
One can imagine a species whose coexistence with others is conditional on the asymmetric relationship between drivers of coexistence and whose populations may become excluded when ATAs erode. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

%future work: study ATAs in environmental variables and how theyre changing - tie in with specialist
Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of the magnitude of ATA contribution and environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in [the marginal/ univariate distributions of] extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

%\underline{multispecies models}
%Need a para about multi-species models - how would things change? One difference is there are more different
%things that might have asymmetric tail association patterns. See the section "General Theory" of Ellner et al 2019 - might
%just want write a short para saying it seems like it should be possible to generalize what we've done along those lines.

@@

The chances of ATAs being important for the coexistence of species increases as more variables are included in the system, 
either through the addition of more species or more drivers (beyond just E and C), or both. 
When multiple species are included in the system, the interaction between different species' E's and C's introduces more relationships 
to consider that may or may not be asymmetrically related and could increase the relative importance of ATA contribution to the species' coexistence. 
Likewise, coexistence between species can also depend on more than two drivers \citep{Ellner_2019}. 
The storage effect decomposition presented here is generalizable to coexistence decompositions of multiple variables.
Following the set up and notation of \cite{Ellner_2019}'s \emph{covariance decomposition}, 
ATA contributions and correlation \emph{per se} can be decomposed from the \emph{associations} between two variables
among many from the appropriate term with parenthesized superscripts with the $\shp$ symbol denoting and separating 
variables that are varying independently from the variables that are covarying whose relationship we are interested in decomposing. 
For example, the term $\Delta^{(XY \shp Z)}$ could be decomposed into $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$. 
As more terms and species are introduced into a system, more relationships between variables must be evaluated,
more of those relationships may be asymmetrically tail associated, and ATAs may contribute more to the maintenance of 
biodiversity in that system. 
%make three different biv dis that are standard normal copula and then pair with a shuffles third
%use to get rbar and subtract from rbar of XY#Z,

%\underline{caveats and future directions}
%Realism - we usede simple/extreme ATA, simple model and simple lab system, futurw work needed to assess 
%how mcuh ATA are important in complex systems, field systems
%what systems? what env vars?
%Ellner et al (2016) points out that coexistence is also mediated by other factors that competition: shared enemies, 
%mutualist, facilitation etc. (these are the exact factors Ellner lists), so future work could explore 
%the importance of ATA for coexistence mediated by these factors.
%Something about spatial mechanisms of coexistence? We considered only mechanisms that applied in a model with
%solely temporal variation, but there are also spatial mechanisms. Need to learn about those and think about
%how to say something about how future work may be able to explore the importance of ATA in that context?

@@

\begin{figure}
\includegraphics[width=.75\textwidth]{../results_figs/fig2.pdf}
\caption{Our simulated lottery-model populations influenced by asymmetric tail associations (ATAs) recovered 
from low densities less often and 
spent more time in a state of one-species dominance, demonstrating that ATAs can influence aspects of 
species coexistence. Panel (a) shows species 1 from the lottery 
model subject to environmental noise with symmetric tail associations (Fig. \ref{fig:noise}b). Panel (b) shows species 1 from the 
same model with the same parameters, but subject to noise with left-tail association (Fig. \ref{fig:noise}a).
Both simulations were run for \Sexpr{M} time steps, though only the first $1000$ are shown. Periods of 
``noticeable coexistence'' of the species, defined to be periods when each species comprises at least $5\%$ of 
total abundance, are marked in grey. The average duration of periods of noticeable coexistence 
was \Sexpr{fig2$coMets$Sym$co} time steps for panel (a) and \Sexpr{fig2$coMets$Asy$co} for panel (b), over the full length of the simulations; 
whereas the average duration of periods of
one-species dominance was \Sexpr{fig2$coMets$Sym$dom} time steps for panel (a) and \Sexpr{fig2$coMets$Asy$dom} for (b). Thus, while periods of 
noticeable coexistence were of similar duration for the two scenarios, periods of one-species dominance were much longer,
on average, for the ATA case. Parameters were $\mu_1 = \Sexpr{fig2$params[1]}$,
$\mu_2 = \Sexpr{fig2$params[2]}$, $\sigma = \Sexpr{fig2$params[3]}$, and $\delta = \Sexpr{fig2$params[4]}$. 
See Introduction and Fig. \ref{fig:noise} caption for model details. \label{fig:pedagog}}
\end{figure}
